An Innovative Storage Stack Addressing Extreme Scale Platforms and Big Data Applications

### 2014-05-02 paper.tex
Lots of editing

Below the user API is an IO forwarding layer that redirects IO calls from the
compute nodes to the IO dispatching layer (in black).  This IO forwarding layer
is analogous to the function of the IO nodes in a BlueGene machine or the
passive data staging processes demonstrated
previously~\cite{nisar:2008:staging,Abbasi:2009:datatap}. The next two layers
have considerable functionality. The IO dispatcher (IOD) serves as the primary
storage interface for the IO stack and is the only way to reach the persistent
storage array in lower layers (in green). Ideally, the IOD layer's
functionality can be optional based on available hardware. Much of the
functionality offered at this layer would shift either up or down the stack as
discussed in detail below. The Distributed Asynchronous Object Storage (DAOS)
layer serves as the persistent storage interface and is intended to be the
traditional file system-like foundation on which everything else is built with
no dependence on any technologies specified above it (in pink and yellow). For
example, the IOD layer with or without burst buffers is not required for DAOS
to operate properly.  Instead, the DAOS layer can handle all of the IO
operations from the user API layer, albiet with the potential performance
penalty of manipulating the shared, persistent storage array. At the bottom is
the Versioning Object Storage Device (VOSD) (in purple).  It serves as the
interface for storing objects of all types efficiently. Think of this layer as
the physical disk interface layer. In terms of Lustre, this would replace the
file system on individual storage targets with an interface friendlier to the
containers of objects concept used in the higher layers.

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product.  These ideas are presented to give a
fuller picture of where the project is going rather than dwelling on any
limitations of the published documents.

@@

The major previous work is really limited to full file systems rather than the
mountain of file system adjustments made over the years.

@@

Other file systems, like GoogleFS~\cite{ghemawat:googlefs}, address distributed
rather than parallel computing and cannot be compared directly. The primary
difference between distributed and parallel file systems is the ability of the
file system to retrieve data simultaneously from multiple clients, in parallel,
and treat the resulting collection of pieces as a single object. Distributed
file systems rely on a single client creating a file, but distributing the set
of files across a wide array of storage devices. The other, popular distributed
file system of note is HDFS~\cite{Shvachko:2010:hdfs} that is distributed as
part of Hadoop. These other file systems are mainly of interest in the context
of the ACG features of FFSIO and will be discussed more in
Section~\ref{sec:acg}.

\section{End-User API Layer}
\label{sec:end-user}
Since the proposal specifies a high-level IO API will be the primary end-user
interface for programmatically interacting with the FFSIO stack, the team used
the HDF-5 API and leverated the Virtual Object Layer (VOL) for the initial
design and implementation demonstration. The VOL provides a way to intercept
HDF-5 API calls just below the API layer to implement alternative
functionality. For this project, rather than the default writing to disk in the
HDF-5 format, the VOL layer is used to interact with the IOD layer and the new
concepts it offers, such as transactions, completely hidden from the user.
This both offers a straightforward way for existing applications to test
against the FFSIO stack and to ensure proper semantics are maintained at the
lower levels. Some additional API functions have been introduced to support
new functionality, such as the ACG support for Hadoop-style applications.  The
additions comprise (Figure~\ref{fig:vol-arch}):

@@

\item
  Object-storage API based on HDF5 to support high-level data models.
  This exposes asynchronous, transactional semantics to the application,
  as well as end-to-end data integrity. It will also allows the usage of
  pointer data types, passing of hints down to the storage and support
  for asynchronous index building, maintenance and querying.

\item
  Virtual Object Layer (VOL) plugin that translates HDF5 API requests
  from applications to IOD API calls.

\item\label{item:fs}
  Function shipping from Compute Nodes (CN) to IO Nodes (ION). This provides
  the application developer with the capacity of sending computation down to the
  IONs and get back results.

@@

HDF5~\cite{hdf5} is a versatile data model containing complex data objects
and metadata. Its information set is a collection of datasets, groups,
datatypes and metadata objects. The data model defines mechanisms for
creating associations between various information items. The main
components of HDF5 are described below.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/hdf5.png}
\caption{Sample file illustrating the hierarchical structure of HDF5 files}
\label{fig:hdf5-sample}
\end{figure}

@@

\item
  \textbf{Dataset}: HDF5 datasets are objects that represent actual data
  or content. Datasets are arrays which can have multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions), and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.

\item
  \textbf{Group}: A group is an explicit association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets or datatypes within it.

@@

The Fast Forward project is working on the adaptation of the HDF5 libraries in
order to expose the storage stack to applications.

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.20]{images/vol-arch.png}
\caption{Architectural view of the VOL abstraction mechanism}
\label{fig:vol-arch}
\end{figure}

\subsection{IOD VOL plugin}
\label{iod-vol-plugin}

The IOD VOL plugin serves as the bridge between HDF5 and the IOD API (Figure \ref{fig:vol-arch}). The application calls the HDF5
library while running on the system's compute nodes. Using the VOL
architecture, the IOD VOL plugin uses a function shipper (RPC library)
to forward the VOL calls to a server component running on the I/O nodes
(IONs). At that point, the VOL calls are translated into I/O Dispatcher
(IOD) API calls and executed at the IONs.

The mapping between HDF5 objects and IOD objects is the following:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  file = container
\item
  dataset = array
\item
  group = key-value store
\item
  attribute = key-value store
\end{itemize}

@@

The IO Forwarding layer offers a mechanism to reduce the concurrency impact of
the massive process count on the storage stack. A current trend mixing MPI with
a threading library like OpenMP is addressing the same issue. In this case,
the storage stack has far fewer end-points in which to receive requests and
data from compute processes. By reducing the number of simultaneous requests,
delays can be reduced. This has been demonstrated for the file open operation
with Lustre~\cite{lofstead:2009:adaptable} and to some degree for accessing the
storage devices themselves~\cite{lofstead:2010:io-variability}. The BlueGene
platform incorporated dedicated hardware to perform this role. The proposed
functionality for this layer has no special features beyond those previously
demonstrated.

@@

The core idea for IOD is to provide a way to manage the IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as data rearrangement, can be moved to the IOD layer
offloading the communication load from the compute nodes. IOD has three main
purposes. First, if the optional burst buffer is available, it works as a fast
cache absorbing write operations for the slower trickle out to the central
storage array. It can also be used to retrieve objects from the central storage
array for more efficient read operations and offers data filtering to make
client reads more efficient.  Second, it offers the transaction mechanism for
controlling data set visibility and to manage faults that would prevent a data
set from being used. Third, data processing operations can be placed in the
IOD. These operations are intended to offer data rearrangement, filtering, and
similar operations prior to data reaching the central storage array.

@@

\subsection{I/O Dispatcher API}
\label{sec:fast-forwards-io-dispatcher-api}

The Fast Forward I/O and Storage project is aimed at integrating many of
the services existing currently in middleware libraries in a
next-generation storage stack, exposed through a non-POSIX,
transactional, asynchronous, object-based API.

From a high-level point of view, the Fast Forward's I/O Dispatcher (IOD)
interface allows the user to:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  transactions
\item
  asynchrony
\item
  objects
\item
  layout
\item
  placement
\item
  formats
\end{itemize}

Type of objects:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  container
\item
  key-value store
\item
  blobs
\item
  multi-dimensional arrays
\end{itemize}

As mentioned previously, IOD allows to store distinct types of objects.
From here on we will limit our discussion to multidimensional arrays
(HDF5 datasets) and leave the treatment of other object types as future
work. A multidimensional array can be sharded among the distinct I/O
nodes in the staging area. The IOD API supports the following
strategies:

@@

It is possible to request the transformation of an object's physical
layout to other formats, having multiple copies of the same objects in
multiple formats if desired. Also, the user can pre-fetch objects from
the storage cluster into the I/O nodes or read them directly from the
storage cluster. At the semantic level (HDF5), indexes can be created
for datasets, which results in being able to read through an index
instead of directly from the base array.

All these distinct alternatives result in having many different ways of
executing the same analysis task. In the future, we plan to extend our
design to incorporate an optimization phase that will prepare a task by
taking into account the distinct possible ways of executing and select
the most promising one. In our subsequent discussions, we consider only
data-movement optimization, i.e., sending the analysis code as close as
possible to the data. In practice, this means we focus on identifying
sharding of datasets and execute code accordingly over the appropriate
shards.

@@

\subsection{I/O Nodes}
The ``I/O nodes'' handle requests forwarded by the scientific applications,
potentially integrate a tier of solid-state devices to absorb the burst of
random operations, and organize/re-format the data so that transfers to/from
the staging area from/to the traditional parallel file system can be done more
efficiently. It also has the capacity to executing analysis on the fresh data
that has just been generated by simulation applications running at the compute
nodes. As the data arrives, re-organization and data preparation can be applied
in order to anticipate the execution of analytical tasks.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-arch.png}
\caption{Exascale Architecture}
\label{fig:exa-arch}
\end{figure}

A common cluster configuration for this type of deployment is shown in
Figure~\ref{fig:exa-arch}. The designated I/O nodes (IONs) are connected to the
compute nodes (CNs) through the same fast fabric (e.g., InfiniBand), while the
connection to the external storage cluster is through a secondary slower
channel (e.g., 10Gb Ethernet). I/O staging thus introduces a new layer in the
application-to-storage stack, as shown in Figure~\ref{fig:exa-stack} (left).

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-stack.png}
\caption{Top-to-bottom view of the component stack}
\label{fig:exa-stack}
\end{figure}

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated as part of the IO
stack~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
The current Fast Forward IOD design recommends incorporating SSDs, but
specifically lists these devices as optional. Unfortunately, not incorporating
burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
the IOD design currently is written assuming burst buffers. This means that the
bulk of IO operations will only hit the IOD layer and proposed functionality,
such as the function shipping, do not discuss the impact on the DAOS layer or
the operators themselves should a burst buffer not be available.  Consider the
important functionality of data rearrangement and doing things like changing
the fast array dimension on shared, spinning media.  The design of DAOS assumes
that it will only be involved when persisting a completed transaction and only
for a fraction of the total transactions created.  Transaction frequency at the
IOD layer can be higher since IOD does not ``flatten'' transactions. Flattening
is the process of transferring the set of copy-on-write changes from
transaction/epoch $n$ to transaction $n+m$ into the DAOS layer. The DAOS layer
mirrors the copy-on-write functionality, but stores the differences between
epochs rather than individual transactions. The primary difference is that at
the DAOS layer, data is stored in large, logically contiguous chunks to manage
the metadata load and hopefully improve read performance.

%One of the bigger concerns is the observation that the original data staging
%proposals all used compute nodes while the newer proposals seek not only to
%make them a fixed portion of the IO stack, but also shared across all machine
%users. The PreDatA~\cite{zheng:2010:predata} paper in particular examines the
%potential costs and advantages of where to place operators similar to the IOD
%proposed function shipping. There are two key takeaways from PreDatA. First,
%placement matters.  Depending on the communication intensity vs. computation
%intensity, where along the IO path to place the operation can matter
%significantly. Second, and more importantly, the amount of time spent
%processing for the operators was stretched to the point where it consumed
%nearly all of the time between IO operations. The given ratios of compute
%processes to staging process examined is representative for future extreme
%scale platforms. If anything, the ratios offer more staging processes than IOD
%processes would be available.
%
In the case of the written IOD design, it describes a fixed-sized staging area
that is partitioned on a per-application basis.
%This is unlikely to be useful
%because of the limited compute and communication capacity to spare to perform
%these operations at a bottleneck in the IO path.  The use of a separate data
%staging area intentionally separate from the IO path allows using operators on
%limited resources leaving the IO path clear for strictly data movement. A
%nuance of this design is discussed in the Design Philosophy below.
%
%By concentrating the Burst Buffers and function shipping into the storage
%stack, three problems arise.  First, the amount of network bandwidth, IO
%bandwidth, and compute power consumed for example operations from a single
%application is likely to completely monopolize the IOD processes. Second, if
%space and time partitioning is used instead, the functionality risks being too
%small to be useful. Third, the long-term hardware performance advantage for
%SSDs is questionable. Recent studies have shown that the erase-before-write
%and interference between reading and writing with flash-based SSDs can cause
%severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
%of an optional use SSD layer in the new Trinity machine at Los Alamos will
%offer a test bed to determine how likely these observed problems would affect a
%production extreme scale platform.
%
%Given these features, the optionality and even incorporation of burst buffers
%in the current design should be carefully considered. Much of the advanced, key
%functionality proposed as they are currently designed ultimately relies on the
%existence of burst buffers to work. Further thought about how to have an IOD
%layer both with and without a burst buffer is required before they can be
%considered optional. As the design stands today, they are a required part of
%the IOD layer for proper functioning. Unfortunately, it is not clear that they
%can address the performance concerns they are intended to cover.

@@

The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSD buffers to the IO nodes. The team
does acknowledge the limitations of this design and intend to ultimately focus
on spreading the IOD layer from the IO nodes into the compute area as well.
This is intended both to help address the limitations of the IO bandwidth and
compute capability of these few nodes for data processing, but also to take
advantage of new layers in the storage hierarchy. By incorporating NVRAM into
compute nodes, new options for buffering data prior to being moved to
centralized storage become available and addresses some of the concerns about
SSD performance. For example, including a small amount of Phase Change memory
into many or most compute nodes offers a way to move data outside of both the
compute and IO path for data and communication intensive operations. Other 
projects~\cite{zheng:2010:predata} have suggested this will have value, but the
cost will have to be considered as part of the overall platform budget. This
lessens the impact of some operators while offering additional options for
places to store data.

@@

Another concern that is acknowledged, but no thought has been applied to, is
the requirement that a single IOD process of the set assigned to an application
is the master for any operation.  Should the number of concurrent applications
exceed the available nodes, sharing an IOD process will be required. The
requirements both in terms of scheduling and resource management were cut from
the project due to funding limitations. Since partitioning of the IOD processes
for exclusive use by particular applications is the assumed operating mode,
should insufficient IOD resources be available, either a job could be delayed
or IOD resources could be reallocated from a different process could be
redeployed for use by the new job. Handling resilience concerns for the IOD
processes must also be address.  These sorts of considerations still need to be
made for a full production system.

@@

>>>>

Below the user API is an IO forwarding layer that redirects IO calls from the
compute nodes to the IO dispatching layer (in black).  This IO forwarding layer
is analogous to the function of the IO nodes in a BlueGene machine or the
passive data staging processes demonstrated
previously~\cite{nisar:2008:staging,Abbasi:2009:datatap}. The next two layers
have considerable functionality. The IO dispatcher (IOD) serves as the primary
storage interface for the IO stack and is the only way to reach the persistent
storage array in lower layers (in green). Ideally, the IOD layer's
functionality can be optional based on available hardware and provided on the
IO Nodes (IONs). Much of the functionality offered at this layer would shift
either up or down the stack as discussed in detail below. The Distributed
Asynchronous Object Storage (DAOS) layer serves as the persistent storage
interface and translation layer between the user-visible object model and the
requirements of the underlying storage infrastructure. It is intended to be the
traditional file system-like foundation on which everything else is built with
no dependence on any technologies specified above it (in pink and yellow). For
example, the IOD layer with or without burst buffers is not required for DAOS
to operate properly.  Instead, the DAOS layer can handle all of the IO
operations from the user API layer, albiet with the potential performance
penalty of manipulating the shared, persistent storage array. At the bottom is
the Versioning Object Storage Device (VOSD) (in purple).  It serves as the
interface for storing objects of all types efficiently for each storage device
in the parallel storage array. Think of this layer as the physical disk
interface layer. In terms of Lustre, this would replace the API on individual
storage devices with an interface friendlier to the containers of objects and
transactions/epochs concepts used in the higher layers.

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product.  These ideas are presented to give a
fuller picture of where the project is going rather than dwelling on any
limitations of the published documents. This is most important to illustrate
how different concepts will work across layers since that information is spread
across multiple documents and may lack a cohesive overall view.

@@

The major previous work is really limited to full file systems rather than the
mountain of file system refinements made over the years.

@@

Other file systems, like GoogleFS~\cite{ghemawat:googlefs}, address distributed
rather than parallel computing and cannot be compared directly. The primary
difference between distributed and parallel file systems is the ability of the
file system to store and retrieve data simultaneously from multiple clients, in
parallel, and treat the resulting collection of pieces as a single object.
Distributed file systems rely on a single client creating a file, but
distributing the set of files across a wide array of storage devices. The
other, popular distributed file system of note is
HDFS~\cite{Shvachko:2010:hdfs} that is distributed as part of Hadoop. These
other file systems are mainly of interest in the context of the ACG features of
FFSIO and will be discussed more in Section~\ref{sec:acg}.

\section{End-User API Layer}
\label{sec:end-user}

Since the proposal specifies a high-level IO API will be the primary end-user
interface for programmatically interacting with the FFSIO stack, the team used
the HDF-5 API and leveraged the Virtual Object Layer (VOL) for the initial
design and implementation demonstration. This also serves as a good test
determining what are strictly necessary extensions to an existing IO API to
support the new functionality.  The additional functionality, such as
transactions, can be ignored for legacy implementations, but these applications
will not be able to take advantage of the asynchronous IO support inherent to
the new API.  The additions comprise (Figure~\ref{fig:vol-arch}):

@@

\item
  API extensions to support new functionality provided by the FFSIO project.
  This includes calls for managing asynchronous request lists, performing
  asynchronous operations, creating and managing transactions, end-to-end data
  integrity, and data type and functions to support the ACG functionality more
  efficiently than the current API.

%\item
%  Virtual Object Layer (VOL) offers the ability to plug in new implementations
%  for HDF-5 semantics. In this case, the HDF-5 calls are implemented using the
%  IOD or DAOS API rather than MPI-IO or POSIX.

\item\label{item:fs}
  Function shipping from Compute Nodes (CN) to IO Nodes (ION). This provides
  the application developer with the capability of sending computation down to
  the IONs and get back results and perform other operations such as indexing
  and data reorganization for more efficient retrieval.

@@

HDF5~\cite{hdf5} has a versatile data model offering complex data objects
and metadata. Its information set is a collection of datasets, groups,
datatypes and metadata objects. The data model defines mechanisms for
creating associations between various information items. The main
components of HDF5 are described below.

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/hdf5.png}
\caption{Sample file illustrating the hierarchical structure of HDF5 files}
\label{fig:hdf5-sample}
\end{figure}

@@

\item
  \textbf{Group}: A group is an explicit association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets, datatypes or attributes within
  it.

\item
  \textbf{Dataset}: HDF5 datasets are objects that represent actual data
  or content. Datasets are arrays with potentially multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions), and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.

@@

@@

For this project, rather than the default writing to disk in the HDF-5 format,
the VOL layer is used to interact with the IOD layer and the different concepts
it offers without requiring all of the functionality be exposed to users. For
example, the containers and objects concept is transparently mapped to the
files and datasets existing HDF5 users are familiar with.  This reduces the
difficulty in porting applications to use the new IO stack.

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/vol-arch.png}
\caption{Architectural view of the VOL abstraction mechanism}
\label{fig:vol-arch}
\end{figure}

%\subsection{IOD VOL plugin}
%\label{iod-vol-plugin}

The IOD VOL plugin serves as the bridge between HDF5 and the IOD or DAOS Layer
(Figure~\ref{fig:vol-arch}). The application calls the HDF5 library while
running on the system's compute nodes. Using the VOL architecture, the IOD VOL
plugin uses a function shipper (RPC library) to forward the VOL calls to a
server component running on the IO nodes (IONs). This function shipping is the
IO Forwarding Layer discussed briefly in Section~\ref{sec:iof}. Once the calls
arrive at the IO nodes, they are translated into I/O Dispatcher (IOD) API calls
and executed at the IONs.

\subsection{HDF5 to FFSIO Mapping}
\label{sec:hdf-to-ffsio}

Since HDF5 has traditionally offered an interface focused on files and the
internal data types, such as datasets, these concepts must be mapped onto the
proposed FFSIO data storage concepts. This mapping is shown in Table~\ref{tab:mapping}

\begin{table}[ht]
    %\vspace{-0.15in}
    \centering
    \caption[HDF5 to FFSIO Mapping]{HDF5 Data Model to FFSIO Data Model Mapping}
    \bigskip
    %\vspace{-0.15in}

    \begin{tabular}{|r|r|}
\hline
HDF5 & FFSIO\\
\hline
file & container \\
dataset & array \\
group & key-value store \\
attribute(s) & key-value store \\
\hline
    \end{tabular}
    \label{tab:mapping}
\end{table}

In Section~\ref{sec:iod} below, the FFSIO types are described in more detail.

@@

The IO Forwarding layer offers a mechanism to reduce the concurrency impact of
the massive process count on the storage stack. A current trend mixing MPI with
a threading library like OpenMP is addressing the same issue. In this case,
the storage stack has far fewer end-points in which to receive requests and
data from compute processes. By reducing the number of simultaneous requests,
delays can be reduced. This has been demonstrated for the file open operation
with Lustre~\cite{lofstead:2009:adaptable} and to some degree for accessing the
storage devices themselves~\cite{lofstead:2010:io-variability}. The BlueGene
platform incorporated dedicated hardware to perform this role. The proposed
functionality for this layer, beyond managing the number of connections to
the IOD layer, is to implement function shipping from the compute nodes to the
IO nodes.

For the basic HDF5 calls, this will work the same as how the Nessie
staging~\cite{lofstead:2011:nessie-staging} shifted the collective IO data
rearrangement calls to a reduced number of processes. The prototype
implementation will only support accessing functionality already deployed to
the IO nodes through an RPC mechanism. This initial implementation will use
Mercury~\cite{Soumagne:2013:mercury} to access the remote functionality. For
dynamically defined functions, a different system will be required leveraging
something like C-on-demand~\cite{abbasi:2011:c-on-demand} or some other dynaimc
deployment and compilation or an interpreter system.

@@

Strictly speaking, the IO Dispatcher layer and included functionality, such as
burst buffers, is an optional component of the proposed IO stack. All of the
functionality, such as function and analysis shipping, transaction management,
and managing asynchronous data movement can be handled by other portions of the
stack. For an extreme scale platform, the IOD layer will be an essential
pressure relief valve for the underlying persistent storage layer. By making it
optional, the proposed stack can be deployed more easily on smaller clusters or
for those with more constrainted budgets. For simplicity, the rest of this
section will describe a full stack including all of the proposed IOD
components.

The core idea for IOD is to provide a way to manage the IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as data rearrangement, can be moved to the IOD layer
offloading the communication load from the compute nodes. IOD has three main
purposes. First, the burst buffers works as a fast cache absorbing write
operations for the slower trickle out to the central storage array. It can also
be used to retrieve objects from the central storage array for more efficient
read operations and offers data filtering to make client reads more efficient.
Second, it offers the transaction mechanism for controlling data set visibility
and to manage faults that would prevent a data set from being used. These
transactions are local to the IOD layer until persisted to the DAOS layer
eliminating the need for burdening the persistent storage with transient data.
Third, data processing operations can be placed in the IOD. These operations
are intended to offer data rearrangement, filtering, and similar operations
prior to data reaching the central storage array.

@@

%\subsection{I/O Dispatcher API}
%\label{sec:fast-forwards-io-dispatcher-api}
%
%The Fast Forward I/O and Storage project is aimed at integrating many of
%the services existing currently in middleware libraries in a
%next-generation storage stack, exposed through a non-POSIX,
%transactional, asynchronous, object-based API.
%
%From a high-level point of view, the Fast Forward's I/O Dispatcher (IOD)
%interface allows the user to:
%
%\begin{itemize}
%\itemsep1pt\parskip0pt\parsep0pt
%\item
%  transactions
%\item
%  asynchrony
%\item
%  objects
%\item
%  layout
%\item
%  placement
%\item
%  formats
%\end{itemize}
%
%Type of objects:

\subsection{FFSIO Data Model Types}
\label{sec:data-model}

With the shift from a directories and stream-of-bytes files model to the
container and object model, some description is required to better understand
how these concepts are being used as well as the raw benefits.

\subsubsection{Container}
As mentioned above, the concept of a container is similar to that of a file in
a traditional file system. However, rather than being in a directory structure,
each container essentially is stored in a hash-space allowing direct access to
any container without regard to the current organizational context of the
file system. For example, there is no need to navigate a directory hierarchy
to name a particular container.

Functionally, a container plays the same role as a file in that it holds a
collection of presumably related data intended to be accessed and manipulated
as a unit. Since this is extended from HDF5 files, the container could also be
viewed as a directory tree of objects where each directory entry specifies
either a sub-directory (group) or some data or attribute.

\subsubsection{Key-Value Store}
This is the base type for the container. Since the container represents
something akin to HDF5 files, everything is stored within a hierarchical
namespace. The root namespace is represented by the base key-value store and
contains a list of all of the objects for this portion of the namespace as well
as additional key-value objects representing sub-groups for the hierarchy. Each
of those key-value store objects works identically.

Attributes are stored in a key-value store object, but use the
multi-dimnensional array and blob objects to store the values for the
attributes.

\subsubsection{Multi-Dimensional Array}
By treating arrays as a special case separate from blobs, additional
opportunities are enabled. For example, by knowing that an object is an array,
proper slicing of that array onto IO nodes can be done without involving higher
levels of the IO stack.

\subsubsection{Blob}
All other data is stored as a stream-of-bytes without regard to the actual
data type.

\subsection{Multi-Dimensional Array Data Distribution}

For both IO performance and to aid in analysis and other data processing, the
multi-dimensional array object can be split across multiple IO nodes. Each
piece of this array is called a {\em shard}. The IOD API supports the following
sharding strategies:

@@

It is possible to request the transformation of an object's physical
layout to other formats, having multiple copies of the same objects in
multiple formats if desired. Also, the user can pre-fetch objects from
the storage cluster into the I/O nodes or read them directly from the
storage cluster. At the semantic level (HDF5), indices can be created
for datasets resulting in being able to read through an index instead of
directly from the base array.

All of these distinct alternatives result in having many different ways for
executing the same analysis task.
%In the future, we plan to extend our
%design to incorporate an optimization phase that will prepare a task by
%taking into account the distinct possible ways of executing and select
%the most promising one.
In the subsequent discussions, we consider only
data-movement optimization, i.e., sending the analysis code as close as
possible to the data. In practice, this means we focus on identifying
sharding of datasets and execute code accordingly over the appropriate
shards.

@@

\subsection{I/O Nodes}
IOD processes are hosted on the IO nodes that interface a general compute area
with the storage array. The IO noes handle requests forwarded by the scientific
applications, potentially integrate a tier of solid-state devices to absorb the
burst of random operations, and organize/re-format the data so that transfers
to/from the staging area from/to the traditional parallel file system can be
done more efficiently. It also has the capacity to executing analysis on the
fresh data that has just been generated by simulation applications running at
the compute nodes. As the data arrives, re-organization and data preparation
can be applied in order to anticipate the execution of analytical tasks.

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/exa-arch.png}
\caption{Exascale Architecture}
\label{fig:exa-arch}
\end{figure}

A common cluster configuration for this type of deployment is shown in
Figure~\ref{fig:exa-arch}. The designated IO nodes (IONs) are connected to the
compute nodes (CNs) through the same fast fabric (e.g., InfiniBand), while the
connection to the external storage cluster is through a secondary slower
channel (e.g., 10Gb Ethernet). By providing additional storage, such as SSDs on
the IO nodes, these nodes are capable of better regulating the IO pressure on
the underlying storage array better than simple forwaring gateways. For this
project, using something like SSDs on the IO nodes is termed a {\em Burst
Buffer} and is disucssed below.

@@

%\begin{figure}[htbp]
%\centering
%\includegraphics[width=\columnwidth]{images/exa-stack.png}
%\caption{Top-to-bottom view of the component stack}
%\label{fig:exa-stack}
%\end{figure}

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated into the existing IO stack
architecture~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
%The current Fast Forward IOD design recommends incorporating SSDs, but
%specifically lists these devices as optional. Unfortunately, not incorporating
%burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
%the IOD design currently is written assuming burst buffers. This means that the
%bulk of IO operations will only hit the IOD layer and proposed functionality,
%such as the function shipping, do not discuss the impact on the DAOS layer or
%the operators themselves should a burst buffer not be available.  Consider the
%important functionality of data rearrangement and doing things like changing
%the fast array dimension on shared, spinning media.  The design of DAOS assumes
%that it will only be involved when persisting a completed transaction and only
%for a fraction of the total transactions created.  Transaction frequency at the
%IOD layer can be higher since IOD does not ``flatten'' transactions. Flattening
%is the process of transferring the set of copy-on-write changes from
%transaction/epoch $n$ to transaction $n+m$ into the DAOS layer. The DAOS layer
%mirrors the copy-on-write functionality, but stores the differences between
%epochs rather than individual transactions. The primary difference is that at
%the DAOS layer, data is stored in large, logically contiguous chunks to manage
%the metadata load and hopefully improve read performance.

%One of the bigger concerns is the observation that the original data staging
%proposals all used compute nodes while the newer proposals seek not only to
%make them a fixed portion of the IO stack, but also shared across all machine
%users. The PreDatA~\cite{zheng:2010:predata} paper in particular examines the
%potential costs and advantages of where to place operators similar to the IOD
%proposed function shipping. There are two key takeaways from PreDatA. First,
%placement matters.  Depending on the communication intensity vs. computation
%intensity, where along the IO path to place the operation can matter
%significantly. Second, and more importantly, the amount of time spent
%processing for the operators was stretched to the point where it consumed
%nearly all of the time between IO operations. The given ratios of compute
%processes to staging process examined is representative for future extreme
%scale platforms. If anything, the ratios offer more staging processes than IOD
%processes would be available.
%

In the case of the written IOD design, it describes a fixed-sized staging area
that is partitioned on a per-application basis. As part of an application
being deployed into the platform, each application will be allocated a fixed
number of IO nodes for exclusive use during the application run. This provides
guarantees about how much burst buffer space and processing capability will be
available for the applications.

%This is unlikely to be useful
%because of the limited compute and communication capacity to spare to perform
%these operations at a bottleneck in the IO path.  The use of a separate data
%staging area intentionally separate from the IO path allows using operators on
%limited resources leaving the IO path clear for strictly data movement. A
%nuance of this design is discussed in the Design Philosophy below.

Future work will generalize this model to potentially support dynamic IO node
allocation and potentially examine the possibility of oversubscription. It will
be strictly necessary to consider shared IO nodes for cases where the number
of deployed applications exceeds the number of IO nodes. This is also under
consideration for future work. This first pass is focused on extreme scale
application runs that use the vast majority of a platform rather than a
capacity cluster where end-to-end performance is a lesser concern.

%By concentrating the Burst Buffers and function shipping into the storage
%stack, three problems arise.  First, the amount of network bandwidth, IO
%bandwidth, and compute power consumed for example operations from a single
%application is likely to completely monopolize the IOD processes. Second, if
%space and time partitioning is used instead, the functionality risks being too
%small to be useful. Third, the long-term hardware performance advantage for
%SSDs is questionable. Recent studies have shown that the erase-before-write
%and interference between reading and writing with flash-based SSDs can cause
%severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
%of an optional use SSD layer in the new Trinity machine at Los Alamos will
%offer a test bed to determine how likely these observed problems would affect a
%production extreme scale platform.
%
%Given these features, the optionality and even incorporation of burst buffers
%in the current design should be carefully considered. Much of the advanced, key
%functionality proposed as they are currently designed ultimately relies on the
%existence of burst buffers to work. Further thought about how to have an IOD
%layer both with and without a burst buffer is required before they can be
%considered optional. As the design stands today, they are a required part of
%the IOD layer for proper functioning. Unfortunately, it is not clear that they
%can address the performance concerns they are intended to cover.

@@

The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSD buffers to the IO nodes. The team
does acknowledge the limitations of this design and intend to ultimately focus
on spreading the IOD layer from the IO nodes into the compute area as well.
This is intended both to help address the limitations of the IO bandwidth and
compute capability of these few nodes for data processing, but also to take
advantage of new layers in the storage hierarchy. By incorporating NVRAM into
compute nodes, new options for buffering data prior to being moved to
centralized storage become available and addresses potential concerns about
SSD performance. For example, including a small amount of Phase Change memory
into many or most compute nodes offers a way to move data outside of both the
compute and IO path for data and communication intensive operations. Other 
projects~\cite{zheng:2010:predata} have suggested this will have value, but the
cost will have to be considered as part of the overall platform budget. This
lessens the impact of some operators while offering additional options for
places to store data.

@@

%Another concern that is acknowledged, but no thought has been applied to, is
%the requirement that a single IOD process of the set assigned to an application
%is the master for any operation.  Should the number of concurrent applications
%exceed the available nodes, sharing an IOD process will be required. The
%requirements both in terms of scheduling and resource management were cut from
%the project due to funding limitations. Since partitioning of the IOD processes
%for exclusive use by particular applications is the assumed operating mode,
%should insufficient IOD resources be available, either a job could be delayed
%or IOD resources could be reallocated from a different process could be
%redeployed for use by the new job. Handling resilience concerns for the IOD
%processes must also be address.  These sorts of considerations still need to be
%made for a full production system.

@@

Overall, the DAOS layer serves as the shared persistent storage interface for
the IO stack. In the case of a data center-wide storage array, the DAOS layer
would be shared across all of the platforms with the upper layers being local
to each individual platform.
